---
title: "Aviation On-Time Performance Analysis"
author: "Joshua Poirier"
date: "Tuesday, January 12, 2016"
output: html_document
---

## Abstract

This study is inspired by the web application and study by Ritchie King and Nate Silver at fivethirtyeight.com titled *Which Flight Will Get You There Fastest?*.  The web application can be found [here](http://projects.fivethirtyeight.com/flights/).  The accompanying [documentation](http://fivethirtyeight.com/features/how-we-found-the-fastest-flights/) describes what they mean by "fastest" flights: 

> Airline A says it will fly you from Seattle to Portland, Oregon, in 45 minutes, but actually takes 60 minutes.  Airline B says it will fly the same route in 75 minutes, and actually takes 70 minutes.  Which flight would you rather take?  That seems easy. Airline A!

I extend this study by considering two cases:  

* **Case 1:** Direct Flight - The consumer wants the fastest flight as defined in the aforementioned study
* **Case 2:** Connecting Flight - The consumer wants an on-time flight in order to avoid missing a tight connection 

I hypothesize that the 3 best and worst airlines for the two cases are different.  To show this, a multivariate regression model will be built for each case to account for confounding variables including **month**, **day of week**, **origin airport**, **destination airport**, and **time of day** in addition to the **airline**.

``` {r loadData, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)

# load ontime data from files in "data" subdirectory
#ontime_data <- do.call(rbind, lapply(paste0("data/", dir("data")), read.csv))
ontime_data <- read.csv("data/2014-01.csv", 
                        colClasses = c(rep("factor", 5), "character", rep("numeric",3)))

# prep the ontime data for analysis
ontime_data$MONTH <- as.factor(month.name[ontime_data$MONTH])
levels(ontime_data$DAY_OF_WEEK) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
ontime_data$CRS_DEP_TIME <- as.numeric(ontime_data$CRS_DEP_TIME)
ontime_data$CRS_DEP_TIME <- cut(ontime_data$CRS_DEP_TIME, breaks=c(-1, 459, 1159, 1759, 2359))
levels(ontime_data$CRS_DEP_TIME) <- c("Evening", "Morning", "Afternoon", "Evening")
ontime_data <- ontime_data[,-10]
bad <- is.na(ontime_data$ARR_DELAY)
ontime_data <- ontime_data[!bad,]

# load airports reference data
airports <- read.csv("reference_data/airports.csv", 
                     colClasses = c("factor", rep("character",2), rep("factor",2), rep("numeric", 2)))

# load airline carriers reference data
carriers <- read.csv("reference_data/carriers.csv",
                     colClasses = c("factor", "character"))

# join the three data sets
ontime_data <- inner_join(ontime_data, airports, by=c("ORIGIN"="iata"))
names(ontime_data)[10:15] <- c("O_AIRPORT", "O_CITY", "O_STATE", "O_COUNTRY", "O_LAT", "O_LONG")
ontime_data <- inner_join(ontime_data, airports, by=c("DEST"="iata"))
names(ontime_data)[16:21] <- c("D_AIRPOT", "D_CITY", "D_STATE", "D_COUNTRY", "D_LAT", "D_LONG")

# now that we have airport coordinates - use simplified formula to calculate target time
ontime_data$TARGET_TIME <- with(ontime_data, 0.117*DISTANCE + 0.517*(O_LONG-D_LONG) + 43.2)
ontime_data$TARGET_DELAY <- with(ontime_data, ACTUAL_ELAPSED_TIME - TARGET_TIME)

# cleanup memory - we don't need airports or carriers data frames anymore
rm(airports, carriers)
```

## Introduction

This study utilizes aviation on-time performance data provided by the United States Department of Transportation [here](http://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236&DB_Short_Name=On-Time).  **Case 2** will use the *ARR_DELAY* field to compare against scheduled arrivals.  **Case 1** will use the difference between the gate-to-gate flight time (*ACTUAL_ELAPSED_TIME* field) and the **target time**.  I calculate target time using the simplified formula shown by Ritchie King and Nate Silver.  

$target time = 0.117 * distance + 0.517 * (lon origin - lon dest) + 43.2$  

This formula produces an estimated travel time in minutes.  **distance** is the Great Circle Distance (shortest distance between two points on the surface of a sphere) and is provided in the data set.  The coefficient **0.117** indicates that flights travel at 513 mph.  **lonorigin** and **londest** represent the longitudes of the originating and destination airports.  30 seconds of flight time is added for every degree of westbound longitude travelled.  The constant **43.2** indicates the time airlines budget for taxiing and inefficient routing (flying around severe weather).

Let's take a look at the first few rows of the data!

```{r previewData, echo=FALSE}
head(ontime_data, 3)
```

## Feature Examination

In this section I take an independent look at the features expected to impact the on-time performance of aircraft.

### Month

### Day of the Week

### Airline

### Origin Airport

### Destination Airport

### Time of Departure



